---
title: "manifoldsR nearest neighbour searches"
vignette: >
  %\VignetteIndexEntry{manifoldsR nearest neighbour searches}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Nearest neighbour searches in manifoldsR

kNN graphs are the core of different manifold learning methods. This vignette
gives you an intro in the provided methods, the advantages and disadvantages
of the different methods. The wrappers can be also used to avoid recomputing
the kNN graph if you wish to test different parameters.

```{r}
#| label: setup
library(manifoldsR)
```

```{r}
#| label: parameters

# number neighbours
k <- 10L
```

## Synthetic data

```{r}
#| label: synthetic data

cluster_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 1000L
)
```

## Exhaustive search

This is simples search type... It uses an exhaustive search index under the
hood with SIMD acceleration. On small data sets, this one might in times be
even faster than the approximate searches we will discuss subsequently. 

```{r}
#| label: exhaustive search

exhaustive <- generate_knn_graph(
  data = cluster_data$data,
  k = k,
  ann_method = "exhaustive"
)

exhaustive
```

There is a set of getters for these classes, see below. An important note is 
that the indices are 0-based (from Rust), so if you wish to use them in R, you
need to be aware of this.

```{r}
#| label: NearestNeigbours getters - indices

# flat representation of the indices
indices_flat <- get_idx_flat(exhaustive)

# as N x idx nearest neigbours matrix
indices_mat <- get_idx_mat(exhaustive)
```

The distances can also be extracted:

```{r}
#| label: NearestNeigbours getters - distances

# flat representation of the distances
dist_flat <- get_dist_flat(exhaustive)

# as N x idx nearest neigbours matrix
dist_mat <- get_dist_mat(exhaustive)
```