---
title: "Nearest neighbour searches"
vignette: >
  %\VignetteIndexEntry{Nearest neighbour searches}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

# Nearest neighbour searches in manifoldsR

kNN graphs are the core of different manifold learning methods. This vignette
gives you an intro in the provided methods, plus the advantages and 
disadvantages of the different methods. The wrappers can be also used to avoid 
recomputing the kNN graph if you wish to test different parameters.

```{r}
#| label: setup
library(manifoldsR)
```

```{r}
#| label: parameters

# number neighbours
k <- 10L

# number samples
n_samples <- 1000L
```

## Synthetic data

We will generate some synthetic data and test the different methods out.

```{r}
#| label: synthetic data

cluster_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = n_samples
)
```

## Exhaustive search

This is simplest search type... It uses an exhaustive search index under the
hood with SIMD acceleration over a flat structure for cache locality. 
On small data sets, this one might in times be even faster than the approximate 
searches we will discuss subsequently. 

```{r}
#| label: exhaustive search

exhaustive <- generate_knn_graph(
  data = cluster_data$data,
  k = k,
  knn_method = "exhaustive"
)

exhaustive
```

There is a set of getters for these classes, see below. An important note is 
that the indices are *0-based (from Rust)*, so if you wish to use them in R, you
need to be aware of this!

```{r}
#| label: NearestNeigbours getters - indices

# flat representation of the indices
indices_flat <- get_idx_flat(exhaustive)

# as N x idx nearest neigbours matrix
indices_mat <- get_idx_mat(exhaustive)
```

The distances can also be extracted easily:

```{r}
#| label: NearestNeigbours getters - distances

# flat representation of the distances
dist_flat <- get_dist_flat(exhaustive)

# as N x idx nearest neigbours matrix
dist_mat <- get_dist_mat(exhaustive)
```

## Approximate searches

This package is designed to deal with large datasets and leverages under the
hood the [`ann-search-rs` crate](https://github.com/GregorLueg/ann-search-rs)
which provides various approximate nearest neighbour searches. The 4 methods
exposed in this package are (+ the exhaustive version above):

| Method | Features | Class | Use case |
|--------|----------|-------|----------|
| BallTree | Simple method, good for small data sets (≤ 100,000) | Tree-based | Small data sets, small dimensionality
| Annoy | Classic in single cell, scales well to data sets of up to 500,000 | Tree-based | Large data sets, small dimensionality
| HNSW | Powerful on large datasets (≥ 500,000), slightly slower index build type, fast queries | Graph-based | Large data sets, small to large dimensionality
| NNDescent | Another graph-based version, good all-rounder when certain sizes are reached (≥ 100,000) | Graph-based | Medium data sets, small to large dimensionality
| IVF | A fast cluster-based index powering libraries like FAISS. | Cluster-based | Medium data sets, small to large dimensionality. You can squeeze out aggressively speed with the n_probe parameter if need be.

manifoldsR defaults to `Balltree`, but you can play around with the parameters & 
methods. The package was written with massive scale in mind, but if you are 
looking at smaller data sets (like in the vignettes), `BallTree` does the job 
(fast). If you have high dimensional data and/or massive amounts of samples,
some of the other methods will shine.

### Speed

Let's just have a look on how they perform on different sizes

```{r}
#| label: approximate nearest neighbour searches - benchmark 1000

microbenchmark::microbenchmark(
  exhaustive = {
    generate_knn_graph(
      data = cluster_data$data,
      k = k,
      knn_method = "exhaustive",
      .verbose = FALSE
    )
  },
  annoy = {
    generate_knn_graph(
      data = cluster_data$data,
      k = k,
      knn_method = "annoy",
      .verbose = FALSE
    )
  },
  hnsw = {
    generate_knn_graph(
      data = cluster_data$data,
      k = k,
      knn_method = "hnsw",
      .verbose = FALSE
    )
  },
  balltree = {
    generate_knn_graph(
      data = cluster_data$data,
      k = k,
      knn_method = "balltree",
      .verbose = FALSE
    )
  },
  nndescent = {
    generate_knn_graph(
      data = cluster_data$data,
      k = k,
      knn_method = "nndescent",
      .verbose = FALSE
    )
  },
  ivf = {
    generate_knn_graph(
      data = cluster_data$data,
      k = k,
      knn_method = "ivf",
      nn_params = params_nn(
        n_list = 3L,
        n_probes = 2L
      ),
      .verbose = FALSE
    )
  },
  times = 1L # single comparison for speed
)
```

On small datasets, the exhaustive search beats everything. We need to build
the indices in the other methods, a cost that only makes sense when we have
sufficient samples to do this. Otherwise, modern CPUs with SIMD can do the
necessary `O(N^2)` in no time on smaller datasets. What about 50k cells?

```{r}
#| label: approximate nearest neighbour searches - benchmark 50k

benchmark_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 50000L
)

microbenchmark::microbenchmark(
  exhaustive = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "exhaustive",
      .verbose = FALSE
    )
  },
  annoy = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "annoy",
      .verbose = FALSE
    )
  },
  hnsw = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "hnsw",
      .verbose = FALSE
    )
  },
  balltree = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "balltree",
      .verbose = FALSE
    )
  },
  nndescent = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "nndescent",
      .verbose = FALSE
    )
  },
  ivf = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "ivf",
      .verbose = FALSE
    )
  },
  times = 1L # single comparison for speed
)
```

We start observing the first pattern, that the approximate nearest neighbour
searches are faster than the exhaustive search. The delta becomes more 
pronounced with more samples:

```{r}
#| label: approximate nearest neighbour searches - benchmark 100k

benchmark_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 100000L
)

# this will take some time...
microbenchmark::microbenchmark(
  exhaustive = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "exhaustive",
      .verbose = FALSE
    )
  },
  annoy = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "annoy",
      .verbose = FALSE
    )
  },
  hnsw = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "hnsw",
      .verbose = FALSE
    )
  },
  balltree = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "balltree",
      .verbose = FALSE
    )
  },
  nndescent = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "nndescent",
      .verbose = FALSE
    )
  },
  ivf = {
    generate_knn_graph(
      data = benchmark_data$data,
      k = k,
      knn_method = "ivf",
      .verbose = FALSE
    )
  },
  times = 1L # single comparison for speed
)
```

### Precision

In the end, these are approximate methods. So, we need to understand how good
they are at recovering the true neighbours. A typical metric is `Recall@k` 
neighbours. Let's just have a look at the smaller data sets here:

```{r}
#| label: approximate nearest neighbour searches - recall

samples <- 25000L

benchmark_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = samples
)

exhaustive <- generate_knn_graph(
  data = benchmark_data$data,
  k = k,
  knn_method = "exhaustive"
)

indices <- c("hnsw", "annoy", "nndescent", "balltree", "ivf")

for (idx in indices) {
  idx_i <- generate_knn_graph(
    data = benchmark_data$data,
    k = k,
    knn_method = idx,
    .verbose = FALSE
  )
  
  recall_i <- sum(get_idx_flat(exhaustive) == get_idx_flat(idx_i)) / 
    (k * samples)

  print(
    sprintf("ANN method %s achieves a Recall of %.3f.", idx, recall_i)
  )
}
```

As you can see, all of these methods are able to (mostly) identify the right 
neighbours (in the right order!). If you wish to understand more how these 
methods work, how they compare, etc., please check this 
[out](https://github.com/GregorLueg/ann-search-rs).