---
title: "Using PHATE"
vignette: >
  %\VignetteIndexEntry{Using PHATE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

# PHATE in manifoldsR

`manifoldsR` provides a fast Rust-based implementation of
[PHATE](https://www.nature.com/articles/s41587-019-0336-3)
(Potential of Heat-diffusion for Affinity-based Trajectory Embedding), a
dimensionality reduction method specifically designed for recovering continuous
structure and trajectories in high-dimensional data.

```{r}
#| label: setup
library(manifoldsR)
library(magrittr)
library(ggplot2)
library(data.table)
```

## Intro

PHATE builds an embedding through a sequence of steps. First, it constructs a
kernel-based affinity graph over the data, using an adaptive bandwidth kernel
controlled by the `decay` parameter. The kernel is then row-normalised to form
a Markov transition matrix — i.e. a diffusion operator — which models random
walks across the data manifold. Diffusion is run for `t` steps, either chosen
automatically via Von Neumann entropy (up to `t_max`) or fixed via `t_custom`.
Running diffusion for multiple steps has a denoising effect: local noise is
smoothed out, and the global manifold geometry becomes more prominent.

From the diffused transition probabilities, PHATE computes a **potential
distance**: it takes the log of the diffused probabilities and uses the
difference between them as a geometrically meaningful distance metric. The
`gamma` parameter controls the precise form of this transformation. Finally,
MDS is applied to this potential distance matrix to produce the low-dimensional
embedding.

Compared to t-SNE and UMAP, PHATE has a distinct character:

- **Designed for continuous structure.** PHATE was developed with biological
  trajectory data in mind. It tends to preserve the topology of branching and
  continuous manifolds far better than t-SNE or UMAP, which tend to fragment
  them into disconnected islands.
- **Inter-point distances are more meaningful.** The diffusion potential gives
  PHATE a more interpretable distance metric than either t-SNE or UMAP. Relative
  distances in the embedding do carry some geometric meaning, though should
  still not be over-interpreted.
- **Fails on genuinely discrete cluster structure.** When data consists of
  well-separated clusters with no continuous geometry connecting them, PHATE
  will attempt to find a manifold that is not really there. The result is that
  clusters get expanded and contorted rather than compactly separated, making
  PHATE a poor choice for such data.
- **The diffusion time matters.** Too little diffusion and noise dominates; too
  much and fine structure is washed out. Automatic selection via entropy knee
  detection works well in practice, but it is worth inspecting whether the
  chosen `t` is sensible for your data.

## Running PHATE

We use the same synthetic datasets as in the UMAP and t-SNE vignette for direct
comparison.

```{r}
#| label: synthetic data

cluster_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 10000L
)

swissrole_data <- manifold_synthetic_data(
  type = "swiss_role",
  n_samples = 1000L
)

trajectory_data <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 10000L
)
```

### Clustered data

PHATE is not designed for discrete cluster structure. When applied to
well-separated clusters, the diffusion operator will attempt to connect them
— expanding and distorting each cluster in the process. This is the expected
failure mode, and it is worth seeing what it looks like. Similar to the others,
let's start first with PCA:

```{r}
#| label: clustered data - pca

pca_clusters <- prcomp(cluster_data$data)

pca_clusters_df <- as.data.table(pca_clusters$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = pca_clusters_df,
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(mapping = aes(colour = cluster), size = 0.75, alpha = 0.5) +
  theme_bw() +
  ggtitle("PCA on clustered data")
```

And now check PHATE:

```{r}
#| label: clustered data - phate

phate_clusters <- phate(
  data = cluster_data$data,
  k = 15L,
  seed = 42L
)

phate_clusters_df <- as.data.table(phate_clusters) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = phate_clusters_df,
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(mapping = aes(colour = cluster), size = 0.75, alpha = 0.5) +
  theme_bw() +
  ggtitle("PHATE on clustered data")
```

As expected, the clusters explode outward and are condensed into nearly single 
points. PHATE is trying to lay out a manifold that does not exist in this data. 
If your data looks like this, use t-SNE or UMAP instead.

### Swiss role

The swiss role is a continuous manifold where PHATE should perform meaningfully
better than t-SNE or UMAP. The diffusion process captures the smooth geometry
of the roll rather than tearing it apart.

```{r}
#| label: swissrole data - phate

phate_swissrole <- phate(
  data = swissrole_data$data,
  k = 15L,
  knn_method = "exhaustive",
  seed = 42L
)

phate_swissrole_df <- as.data.table(phate_swissrole) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  .[, z_axis := swissrole_data$data[, 3]]

ggplot(
  data = phate_swissrole_df,
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(
    mapping = aes(fill = z_axis),
    shape = 21,
    size = 2,
    alpha = 0.5
  ) +
  scale_fill_viridis_c() +
  theme_bw() +
  ggtitle("PCA on swissrole data")
```

PHATE preserves the continuous colour gradient along the roll substantially
better than t-SNE, and considerably better than UMAP. For manifolds like this,
PHATE is the stronger choice.

### Trajectory

The trajectory data is where PHATE was built to shine. Branching trajectories
are the canonical use case.

```{r}
#| label: trajectory data - phate

phate_trajectory <- phate(
  data = trajectory_data$data,
  k = 15L,
  seed = 42L
)

phate_trajectory_df <- as.data.table(phate_trajectory) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  .[, branch := trajectory_data$membership]

ggplot(
  data = phate_trajectory_df[
    sample(1:nrow(phate_trajectory_df), nrow(phate_trajectory_df)),
  ],
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(
    mapping = aes(colour = branch),
    alpha = 0.5,
    size = 0.75
  ) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo") +
  ggtitle("PHATE on trajectory data")
```

PHATE recovers the branching structure cleanly. The continuous progression
along each branch is preserved, and the branch points are identifiable. This
is the kind of data PHATE was designed for.

## Using pre-computed kNN graphs

As with t-SNE, if you want to explore multiple parameter combinations without
repeating the kNN search, you can precompute the graph and pass it directly.

```{r}
#| label: phate - pre-computed kNN

precomputed_knn <- generate_knn_graph(
  data = trajectory_data$data,
  k = 15L,
  knn_method = "balltree",
  .verbose = FALSE
)

phate_from_knn <- phate(
  data = trajectory_data$data,
  knn = precomputed_knn,
  k = 15L,
  seed = 42L
)

phate_from_knn_df <- as.data.table(phate_from_knn) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  .[, branch := trajectory_data$membership]

ggplot(
  data = phate_from_knn_df[
    sample(1:nrow(phate_from_knn_df), nrow(phate_from_knn_df)),
  ],
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.75) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo")  +
  ggtitle("PHATE from kNN graph (version 1)")
```

With the kNN graph fixed, we can quickly vary PHATE-specific parameters. For
example, trying a fixed diffusion time rather than automatic selection:

```{r}
#| label: phate - pre-computed kNN, different params

phate_from_knn_fixed_t <- phate(
  data = trajectory_data$data,
  knn = precomputed_knn,
  k = 15L,
  phate_params = params_phate(t_custom = 10L),
  seed = 42L
)

phate_from_knn_fixed_t_df <- as.data.table(phate_from_knn_fixed_t) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  .[, branch := trajectory_data$membership]

ggplot(
  data = phate_from_knn_fixed_t_df[
    sample(1:nrow(phate_from_knn_fixed_t_df), nrow(phate_from_knn_fixed_t_df)),
  ],
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.75) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo")  +
  ggtitle("PHATE from kNN graph (version 2)")
```

## Landmark PHATE

For large datasets, computing and storing the full N x N diffusion operator is
prohibitive — both in memory and computation. The landmark approach addresses
this by selecting a small, representative subset of points (landmarks), running
the full diffusion only over those, and then interpolating the remaining points
into the landmark space. The default behaviour is to use `2048L` in this package
(if you wonder why 2048, it's for the SIMD-acceleration in the k-means under the 
hood which work better on 2^x numbers).

The three landmark selection strategies available via `landmark_method` differ
in how they choose these representative points:

- **`"spectral"`** (default): landmarks are selected via spectral clustering on
  the affinity graph. This tends to give the most geometrically representative
  coverage, particularly for data with complex topology, but is the most
  expensive of the three.
- **`"random"`**: landmarks are drawn uniformly at random. Fast, but coverage
  can be uneven in datasets with non-uniform density.
- **`"density"`**: landmarks are drawn from a dampened node degree-weighted 
  distribution. This makes it a bit more precise than pure random, but rare
  branches might be missed.

The number of landmarks is controlled by `n_landmarks`. As a rough guide, 
500–2000 landmarks is sufficient for most datasets up to a few hundred thousand 
points. Too few and the interpolation will miss fine structure; too many and you 
lose most of the speed benefit. If you try running this on large data sets
with `n_landmarks = NULL`, the function will warn you... It will be nasty on
larger data sets as you have to materialise a `N x N` dense matrix, calculated
the potential distances over `N x N`, instead of `N x L`, etc.

```{r}
#| label: phate - landmarks

large_trajectory <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 50000L
)

phate_landmark <- phate(
  data = large_trajectory$data,
  k = 15L,
  phate_params = params_phate(
    n_landmarks = 1024L,
    landmark_method = "spectral"
  ),
  seed = 42L
)

phate_landmark_df <- as.data.table(phate_landmark) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  .[, branch := large_trajectory$membership]

ggplot(
  data = phate_landmark_df[
    sample(1:nrow(phate_landmark_df), nrow(phate_landmark_df)),
  ],
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.75) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo") +
  ggtitle("PHATE with 1024 landmarks (spectral)")
```

We can even further accelerate this with the density or random-based version

```{r}
#| label: phate - landmarks, using random

large_trajectory <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 50000L
)

phate_landmark <- phate(
  data = large_trajectory$data,
  k = 15L,
  knn_method = "balltree",
  phate_params = params_phate(
    n_landmarks = 1024L,
    landmark_method = "random"
  ),
  seed = 42L
)

phate_landmark_df <- as.data.table(phate_landmark) %>%
  `colnames<-`(c("PHATE1", "PHATE2")) %>%
  dplyr::mutate(branch = large_trajectory$membership)

ggplot(
  data = phate_landmark_df[
    sample(1:nrow(phate_landmark_df), nrow(phate_landmark_df)),
  ],
  mapping = aes(x = PHATE1, y = PHATE2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.75) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo") +
  ggtitle("PHATE with 1024 landmarks (random)")
```

We can observe some differences here, but the overall trajectory is correctly
identified.

## Benchmarks

Let's compare `manifoldsR` against `phateR` (the reference R implementation,
itself a wrapper around the Python `phate` library) on the trajectory data.

```{r}
#| label: benchmark
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE) && requireNamespace("phateR", quietly = TRUE) && requireNamespace("reticulate", quietly = TRUE)

local({
  env_name <- "r-manifolds"
  
  if (!reticulate::virtualenv_exists(env_name)) {
    reticulate::virtualenv_create(env_name)
    reticulate::virtualenv_install(
      env_name, 
      packages = c("phate", "umap-learn", "numpy", "scipy")
    )
  }
  
  if (!reticulate::py_available()) {
    reticulate::use_virtualenv(env_name, required = TRUE)
  }
  
  missing_pkgs <- Filter(
    Negate(reticulate::py_module_available),
    c("phate", "umap")
  )
  if (length(missing_pkgs) > 0L) {
    reticulate::virtualenv_install(
      env_name,
      packages = c("phate", "umap-learn", "numpy", "scipy")
    )
  }
})

benchmark_data <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 10000L
)

microbenchmark::microbenchmark(
  phateR = {
    phateR::phate(
      data = benchmark_data$data,
      knn = 5L,
      n.jobs = -1,
      verbose = FALSE
    )
  },
  manifold_phate = {
    phate(
      data = benchmark_data$data,
      k = 5L,
      seed = 42L,
      .verbose = FALSE
    )
  },
  times = 1L
)
```

And on a larger data set with additionally the random landmark version

```{r}
#| label: benchmark - landmarks
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

benchmark_data_large <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 100000L
)

microbenchmark::microbenchmark(
  phateR = {
    phateR::phate(
      data = benchmark_data_large$data,
      knn = 5L,
      npca = 100L,
      n.jobs = -1,
      verbose = FALSE
    )
  },
  manifold_phate_spectral = {
    phate(
      data = benchmark_data_large$data,
      k = 5L,
      knn_method = "nndescent", # good allrounder for larger data sets
      seed = 42L,
      .verbose = FALSE
    )
  },
  manifold_phate_random = {
    phate(
      data = benchmark_data_large$data,
      k = 5L,
      knn_method = "nndescent", # good allrounder for larger data sets
      phate_params = params_phate(landmark_method = "random"),
      seed = 42L,
      .verbose = FALSE
    )
  },
  times = 1L
)
```

As with t-SNE and UMAP, the speed advantage of the Rust implementation comes 
from the entire pipeline — kNN construction, kernel computation, diffusion, and
MDS — running within the Rust kernel without round-tripping through Python in
this case. Usually, the differences become more pronounced with larger data 
sets. 