---
title: "Using t-SNE"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Using t-SNE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

# t-SNE in manifoldsR

`manifoldsR` provides a fast Rust-based implementation of
[t-SNE](https://www.jmlr.org/papers/v9/vandermaaten08a.html)
(t-Distributed Stochastic Neighbour Embedding), a widely used method for
non-linear dimensionality reduction and visualisation. There is additionally
the FFT-accelerated Interpolation-based t-SNE version available, based on
the brilliant work of 
[George, et al.](https://www.nature.com/articles/s41592-018-0308-4). 

```{r}
#| label: setup
library(manifoldsR)
library(magrittr)
library(ggplot2)
library(data.table)
```

## Intro

t-SNE works by modelling pairwise similarities between points in the
high-dimensional space as a Gaussian distribution, and between points in the
low-dimensional embedding as a Student t-distribution. It then minimises the
Kullback-Leibler divergence between these two distributions via gradient
descent. The heavy tails of the t-distribution are the key trick: they allow
dissimilar points to be modelled far apart in the embedding without requiring
enormous gradients, which alleviates the crowding problem that plagues
earlier methods.

Compared to UMAP, t-SNE has a few notable properties worth keeping in mind:

- **Better at preserving local structure in continuous manifolds.** Because
  the loss is symmetric and driven by pairwise probabilities rather than a
  graph connectivity objective, t-SNE tends to tear continuous structures
  apart less aggressively than UMAP. That said, it is not immune to this
  problem, particularly at low perplexity values.
- **Slower.** The naive implementation is `O(N^2)`, which quickly becomes
  prohibitive. The Barnes-Hut approximation (`"bh"`) reduces this to
  `O(N log N)`, and the FFT-accelerated version (`"fft"`) further reduces it
  to `O(N)` — though the FFT variant is currently only supported on Unix
  systems due to cross-compilation constraints with FFTW.
- **No out-of-sample extension.** Unlike UMAP, t-SNE embeddings cannot
  trivially be used to project new data points into the learned space. (To be
  fair, something that is not - yet - support in the package on the UMAP side.)
- **Distances within the embedding are not meaningful.** Like UMAP, t-SNE
  is a visualisation tool. Cluster sizes and inter-cluster distances in the
  embedding should [not be over-interpreted.](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1011288) 

A key hyperparameter is **perplexity**, which loosely controls the effective
number of neighbours each point considers. Typical values are between 5 and 50.
Lower values focus on very local structure; higher values capture more global
patterns. The learning rate and number of epochs also matter, particularly for
large datasets.

## Running t-SNE

We use the same synthetic datasets as in the UMAP vignette for direct
comparison.

```{r}
#| label: synthetic data
cluster_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 25000L
)

swissrole_data <- manifold_synthetic_data(
  type = "swiss_role",
  n_samples = 1000L
)

trajectory_data <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 25000L
)
```

### Clustered data

t-SNE is well known for producing crisp cluster separation. Let's see how it
handles the clustered synthetic data. But first, let's check PCA...

```{r}
#| label: clustered data - pca
pca_clusters <- prcomp(cluster_data$data)

pca_clusters_df <- as.data.table(pca_clusters$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = pca_clusters_df,
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(mapping = aes(colour = cluster), size = 0.75, alpha = 0.5) +
  theme_bw() +
  ggtitle("PCA on cluster data")
```

And now tSNE:

```{r}
#| label: clustered data - tsne
tsne_clusters <- tsne(
  data = cluster_data$data,
  perplexity = 30,
  seed = 42L
)

tsne_clustered_df <- as.data.table(tsne_clusters) %>%
  `colnames<-`(c("tSNE1", "tSNE2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = tsne_clustered_df,
  mapping = aes(x = tSNE1, y = tSNE2)
) +
  geom_point(mapping = aes(colour = cluster), size = 0.75, alpha = 0.5) +
  theme_bw() +
  ggtitle("tSNE (BH) on cluster data")
```

As expected, t-SNE separates the clusters cleanly. Worth noting: the
inter-cluster distances and cluster sizes visible here are not directly
interpretable — t-SNE optimises for local neighbourhood fidelity, not a
faithful global geometry.

### Swissrole

Let's look at the swiss role, where UMAP struggled due to its graph-based
objective. We will need higher perplexity values here to emphasise the global
structure more. Similar to last UMAP, let's start with PCA:

```{r}
#| label: swissrole data - pca
pca_swiss_role <- prcomp(swissrole_data$data)

pca_swiss_role_df <- as.data.table(pca_swiss_role$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  .[, z_axis := swissrole_data$data[, 3]]

ggplot(
  data = pca_swiss_role_df,
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(
    mapping = aes(fill = z_axis),
    shape = 21,
    size = 2,
    alpha = 0.5
  ) +
  scale_fill_viridis_c() +
  theme_bw() +
  ggtitle("PCA on swissrole")
``` 

And compare against tSNE:

```{r}
#| label: swissrole data - tsne
tsne_swissrole <- tsne(
  data = swissrole_data$data,
  knn_method = "exhaustive",
  perplexity = 50L, # we need to increase perplexity to capture the global structure
  seed = 42L
)

tsne_swissrole_df <- as.data.table(tsne_swissrole) %>%
  `colnames<-`(c("tSNE1", "tSNE2")) %>%
  .[, z_axis := swissrole_data$data[, 3]]

ggplot(
  data = tsne_swissrole_df,
  mapping = aes(x = tSNE1, y = tSNE2)
) +
  geom_point(
    mapping = aes(fill = z_axis),
    shape = 21,
    size = 2,
    alpha = 0.5
  ) +
  scale_fill_viridis_c() +
  theme_bw() +
  ggtitle("tSNE (BH) on swissrole")
```

t-SNE handles the swiss role more gracefully than UMAP, preserving more of the
continuous colour gradient along the roll. That said, it is still not perfect
— this is still better approached with PCA or dedicated manifold unrolling
methods (check out PHATE here).

### Trajectory

Lastly, let's look at trajectory type data...

```{r}
#| label: trajectory data - pca
pca_trajectory <- prcomp(trajectory_data$data)

pca_trajectory_df <- as.data.table(pca_trajectory$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  .[, branch := trajectory_data$membership]

# will mix the labels to visualise the continuum better
ggplot(
  data = pca_trajectory_df[
    sample(1:nrow(pca_trajectory_df), nrow(pca_trajectory_df)),
  ],
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.75) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo") +
  ggtitle("PCA on trajectory")
``` 

And tSNE

```{r}
#| label: trajectory data - tsne
tsne_trajectory <- tsne(
  data = trajectory_data$data,
  perplexity = 25,
  seed = 42L
)

tsne_trajectory_df <- as.data.table(tsne_trajectory) %>%
  `colnames<-`(c("tSNE1", "tSNE2")) %>%
  .[, branch := trajectory_data$membership]

ggplot(
  data = tsne_trajectory_df[
    sample(1:nrow(tsne_trajectory_df), nrow(tsne_trajectory_df)),
  ],
  mapping = aes(x = tSNE1, y = tSNE2)
) +
  geom_point(
    mapping = aes(colour = branch),
    alpha = 0.5,
    size = 0.75
  ) +
  theme_bw() +
  scale_colour_viridis_c(option = "turbo") +
  ggtitle("tSNE (BH) on trajectory")
```

t-SNE does a better job of recovering the branching trajectory structure.
Compared to UMAP, the continuous progression along each branch is somewhat
better preserved, though the embedding still introduces distortions that are
not present in a PCA plot and without knowing in detail how the differentiation
path would go, it would not be clear.

## Using pre-computed kNN graphs

If you want to test multiple parameter combinations while keeping the kNN graph
fixed (e.g. varying perplexity or the number of epochs), you can precompute
the kNN graph and pass it in directly.

```{r}
#| label: tsne - pre-computed kNN
precomputed_knn <- generate_knn_graph(
  data = cluster_data$data,
  k = 30L, # we will use 30 neighbours here...
  .verbose = FALSE
)

tsne_from_knn <- tsne(
  data = cluster_data$data,
  knn = precomputed_knn,
  perplexity = 10,
  seed = 42L
)

tsne_from_knn_df <- as.data.table(tsne_from_knn) %>%
  `colnames<-`(c("tSNE1", "tSNE2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = tsne_from_knn_df,
  mapping = aes(x = tSNE1, y = tSNE2)
) +
  geom_point(mapping = aes(colour = cluster), alpha = 0.5, size = 0.75) +
  theme_bw() +
  ggtitle("tSNE on pre-computed kNN (version 1)")
```

Now we can quickly re-run with different t-SNE parameters without repeating the
kNN search:

```{r}
#| label: tsne - pre-computed kNN, different params
tsne_from_knn_highperplex <- tsne(
  data = cluster_data$data,
  knn = precomputed_knn,
  perplexity = 15,
  tsne_params = params_tsne(n_epochs = 300L, lr = 100.0), # we will stop early to show a weirder embedding
  seed = 42L
)

tsne_from_knn_highperplex_df <- as.data.table(tsne_from_knn_highperplex) %>%
  `colnames<-`(c("tSNE1", "tSNE2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = tsne_from_knn_highperplex_df,
  mapping = aes(x = tSNE1, y = tSNE2)
) +
  geom_point(mapping = aes(colour = cluster), alpha = 0.5, size = 0.75) +
  theme_bw() +
  ggtitle("tSNE on pre-computed kNN (version 2)")
```

## FFT-accelerated t-SNE

For large datasets, the Barnes-Hut approximation `O(N log N)` can still be
slow. The FFT-accelerated variant reduces this to `O(N)` by interpolating the
repulsive forces on a grid, making it substantially faster at scale. The
constant however is higher. This version of tSNE becomes interesting in
situations of ≥ 100,000 samples and more.

**Note: the FFT variant is only available on Unix-based systems!**

```{r}
#| label: tsne - fft
#| eval: !expr .Platform$OS.type == "unix"
tsne_fft <- tsne(
  data = cluster_data$data,
  perplexity = 30,
  approx_type = "fft",
  seed = 42L
)

tsne_fft_df <- as.data.table(tsne_fft) %>%
  `colnames<-`(c("tSNE1", "tSNE2")) %>%
  .[, cluster := as.factor(cluster_data$membership)]

ggplot(
  data = tsne_fft_df,
  mapping = aes(x = tSNE1, y = tSNE2)
) +
  geom_point(mapping = aes(colour = cluster), alpha = 0.5, size = 0.75) +
  theme_bw() +
  ggtitle("tSNE (FFT) on cluster data")
```

The embeddings are qualitatively similar to the BH version, with the FFT
version trading a small amount of precision for a significant speed gain on
large datasets.

**Important note:** On large data sets (and the effect grows with N) you may
observe that the clusters do not separate into distinct islands but instead
relax into a single, gapless disc tiled into Voronoi-like cells, with clusters
meeting along shared boundaries rather than separated by empty space. This is
not a convergence failure: the optimiser has reached the equilibrium implied by
its parameters. It arises because the main optimisation phase runs at an
exaggeration of 1, and at high point density the attraction-repulsion balance
at that setting is a space-filling layout. If you observe this set 
`late_exag_factor` to values of 2 to 4 (pending your data set size.)

## Benchmarks

Let's compare the BH implementation in `manifoldsR` against `Rtsne` (the most
widely used t-SNE implementation in R, itself a wrapper around the C++ BH
t-SNE code). The `tsne` package (pure R, naive `O(N^2)` implementation) is
just too slow to be worth comparing here... 

```{r}
#| label: benchmark - small
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE) && requireNamespace("Rtsne", quietly = TRUE)
benchmark_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 5000L
)

microbenchmark::microbenchmark(
  Rtsne = {
    Rtsne::Rtsne(
      X = benchmark_data$data,
      dims = 2,
      perplexity = 30,
      verbose = FALSE
    )
  },
  manifold_bh = {
    tsne(
      data = benchmark_data$data,
      perplexity = 30,
      approx_type = "bh",
      seed = 42L,
      .verbose = FALSE
    )
  },
  times = 1L
)
```

The impact here is massive already. Let's see what happens with BH and FFT? (To
note, FFT's advantage becomes larger the bigger the data set due to its `O(N)`
complexity.)

```{r}
#| label: benchmark - larger
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE) && .Platform$OS.type == "unix"
benchmark_data_large <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 75000L
)

microbenchmark::microbenchmark(
  manifold_bh = {
    tsne(
      data = benchmark_data_large$data,
      perplexity = 30,
      approx_type = "bh",
      seed = 42L,
      .verbose = FALSE
    )
  },
  manifold_fft = {
    tsne(
      data = benchmark_data_large$data,
      perplexity = 30,
      approx_type = "fft",
      seed = 42L,
      .verbose = FALSE
    )
  },
  times = 1L
)
```

The speed advantage of the Rust implementation comes from a combination of
optimised memory layout, aggressive inlining, and the fact that the entire
pipeline — kNN construction, probability computation, and gradient descent —
runs within the Rust kernel without round-tripping through R. If you like
tSNE, but thought it's too slow... we got you.
