---
title: "Using UMAP"
vignette: >
  %\VignetteIndexEntry{Using UMAP}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

# UMAP in manifoldR

`manifoldsR` provides very rapid implementation of 
[UMAP](https://arxiv.org/abs/1802.03426), based on the brilliant work in the
[uwot](https://github.com/jlmelville/uwot) package. The aim of this version
is to use a Rust-based version and accelerate the algorithm even further.

```{r}
#| label: setup
library(manifoldsR)
library(magrittr)
library(ggplot2)
```

## Intro

UMAP works by constructing a weighted k-nearest neighbour graph in the 
high-dimensional space (see the knn_searches vignette for more details here!), 
where edge weights reflect the likelihood that two points are connected given 
their distance under a local, adaptive metric. 
Subsequently, It then minimises a cross-entropy loss between this 
high-dimensional graph and a low-dimensional analogue, effectively pulling 
connected nodes together and pushing unconnected ones apart.
This objective is why **UMAP excels at preserving cluster structure**: if two points 
are unlikely neighbours in high dimensions, the repulsion term drives them apart 
in the embedding, producing clean separation. 
The flip side is that UMAP has 
**no incentive to preserve the relative distances along a continuum**. For 
trajectory or manifold data where the meaningful signal 
is a smooth progression, the loss treats weakly-connected distant points as 
things to repel rather than as part of a coherent structure. The result is that 
continuous manifolds tend to get torn apart into spurious clusters. Distances 
and even topology within an embedding should **therefore be interpreted with caution** — 
[UMAP is a visualisation tool, not a metric space!](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1011288).

## Running UMAP

We are going to generate three different data sets to showcase the strengths
(and weaknesses) of UMAP:

```{r}
#| label: synthetic data

cluster_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 25000L
)

swissrole_data <- manifold_synthetic_data(
  type = "swiss_role",
  n_samples = 1000L
)

trajectory_data <- manifold_synthetic_data(
  type = "trajectory",
  n_samples = 25000L
)
```

### Clustered data

Let's start with something UMAP does well - visualise well separated clusters.
We will first contrast this against standard PCA to have some bases...

```{r}
#| label: clustered data - pca

pca_clusters <- prcomp(cluster_data$data)

pca_clusters_df <- as.data.frame(pca_clusters$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  dplyr::mutate(cluster = as.factor(cluster_data$membership))

ggplot(
  data = pca_clusters_df,
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(mapping = aes(colour = cluster), size = 0.25, alpha = 0.5) +
  theme_bw()
```

PCA does an okay job here; however, several of the clusters are clearly mixed.
The separation likely occurs with different combination of the PCs, but let's
see what UMAP can do here on just two dimensions.

```{r}
#| label: clustered data - umap

# we are going to use the SGD optimiser on this small data set.
# the parallelised Adam optimiser ala uwot::umap2 will be shown later.
umap_clusters <- umap(
  data = cluster_data$data,
  nn_method = "exhaustive",
  umap_params = params_umap(optimiser = "sgd"),
  min_dist = 0.5,
  k = 10L
)

umap_clustered_df <- as.data.frame(umap_clusters) %>%
  `colnames<-`(c("UMAP1", "UMAP2")) %>%
  dplyr::mutate(cluster = as.factor(cluster_data$membership))


ggplot(
  data = umap_clustered_df,
  mapping = aes(x = UMAP1, y = UMAP2)
) +
  geom_point(mapping = aes(colour = cluster), alpha = 0.5, size = 0.25) +
  theme_bw()
```

In this case, UMAP is clearly able to separate out the individual clusters (as
expected). However, in other cases, UMAP does worse... Let's look at the
SwissRole example.

### Swissrole

```{r}
#| label: swissrole data - pca

pca_swiss_role <- prcomp(swissrole_data$data)

pca_swiss_role_df <- as.data.frame(pca_swiss_role$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  dplyr::mutate(z_axis = swissrole_data$data[, 3])

ggplot(
  data = pca_swiss_role_df,
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(
    mapping = aes(fill = z_axis), 
    shape = 21, 
    size = 2, 
    alpha = 0.5
  ) +
  scale_fill_viridis_c() +
  theme_bw()
``` 

PCA clearly captures the Swissrole structure. How does UMAP perform here ... ?

```{r}
#| label: swissrole data - umap

umap_swissrole <- umap(
  data = swissrole_data$data,
  nn_method = "exhaustive",
  umap_params = params_umap(optimiser = "sgd"),
  min_dist = 0.5,
  k = 10L
)

umap_swissrole_df <- as.data.frame(umap_swissrole) %>%
  `colnames<-`(c("UMAP1", "UMAP2")) %>%
  dplyr::mutate(z_axis = swissrole_data$data[, 3])


ggplot(
  data = umap_swissrole_df,
  mapping = aes(x = UMAP1, y = UMAP2)
) +
  geom_point(
    mapping = aes(fill = z_axis), 
    shape = 21, 
    size = 2, 
    alpha = 0.5
  ) +
  scale_fill_viridis_c() +
  theme_bw()
```

UMAP rips apart the swiss role structure. Due to the nature of the loss, it is
not a great algorithm for data with a continuum-type manifold. This is also
apparent in the trajectory type data:

### Trajectory

Let's start with PCA again

```{r}
#| label: trajectory data - pca

pca_trajectory <- prcomp(trajectory_data$data)

pca_trajectory_df <- as.data.frame(pca_trajectory$x[, 1:2]) %>%
  `colnames<-`(c("PC1", "PC2")) %>%
  dplyr::mutate(branch = trajectory_data$membership)

# will mix the labels to visualise the continuum better
ggplot(
  data = pca_trajectory_df[
    sample(1:nrow(pca_trajectory_df), nrow(pca_trajectory_df)), 
  ],
  mapping = aes(x = PC1, y = PC2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.25) +
  theme_bw() +
  scale_colour_viridis_c(option = "mako")
``` 

We can appreciate the branching and the continuum in the data.

```{r}
#| label: trajectory data - umap

umap_trajectory <- umap(
  data = trajectory_data$data,
  nn_method = "exhaustive",
  umap_params = params_umap(optimiser = "sgd"),
  min_dist = 0.5,
  k = 10L
)

umap_trajectory_df <- as.data.frame(umap_trajectory) %>%
  `colnames<-`(c("UMAP1", "UMAP2")) %>%
  dplyr::mutate(branch = trajectory_data$membership)


ggplot(
  data = umap_trajectory_df[
    sample(1:nrow(pca_trajectory_df), nrow(pca_trajectory_df)), 
  ],
  mapping = aes(x = UMAP1, y = UMAP2)
) +
  geom_point(mapping = aes(colour = branch), alpha = 0.5, size = 0.25) +
  theme_bw() +
  scale_colour_viridis_c(option = "mako")
```

While the structure is kinda kept and one can see some type of trajectory, the
clearness of the PCA is lost here.

## UMAP speed and optimisers

The package provides several optimisers with one of them designed for speed,
i.e., the parallelised ADAM optimiser. This one is designed to deal with LARGE
data sets you throw at the problem and it can generate the embeddings very
fast. Let's check out the other optimisers:

```{r}
#| label: umap - sequential adam

umap_clusters <- umap(
  data = cluster_data$data,
  nn_method = "exhaustive",
  umap_params = params_umap(optimiser = "adam"),
  min_dist = 0.35,
  k = 10L
)

umap_clustered_df <- as.data.frame(umap_clusters) %>%
  `colnames<-`(c("UMAP1", "UMAP2")) %>%
  dplyr::mutate(cluster = as.factor(cluster_data$membership))


ggplot(
  data = umap_clustered_df,
  mapping = aes(x = UMAP1, y = UMAP2)
) +
  geom_point(mapping = aes(colour = cluster), alpha = 0.5, size = 0.25) +
  theme_bw()
```

This one is actually a bit slower, but can in times generate nicer embeddings.
But let's look at the fast version... Parallised Adam (the default):

```{r}
#| label: umap - parallel adam

umap_clusters <- umap(
  data = cluster_data$data,
  nn_method = "exhaustive",
  min_dist = 0.35,
  k = 10L
)

umap_clustered_df <- as.data.frame(umap_clusters) %>%
  `colnames<-`(c("UMAP1", "UMAP2")) %>%
  dplyr::mutate(cluster = as.factor(cluster_data$membership))


ggplot(
  data = umap_clustered_df,
  mapping = aes(x = UMAP1, y = UMAP2)
) +
  geom_point(mapping = aes(colour = cluster), alpha = 0.5, size = 0.25) +
  theme_bw()
```

This one is SUBSTANTIALLY faster, as the gradients for each node in the graph
are collected in parallel and then applied. This allows us to make usage of
heavy multi-threading which Rust supports beautifully via Rayon. In terms of 
how does this version compare against other versions, let's do a benchmark:

```{r}
#| label: speed benchmark - small data set
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE) && requireNamespace("umap", quietly = TRUE) && requireNamespace("uwot", quietly = TRUE)

benchmark_data <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 5000L
)

microbenchmark::microbenchmark(
  # umap version in pure R
  umap = {
    umap::umap(
      d = benchmark_data$data,
      method = "naive"
    )
  },
  # uwot - umap, for example used in Seurat
  uwot = {
    uwot::umap(X = benchmark_data$data)
  },
  # uwot - umap2 version with parallel Adam under the hood
  # the approach that was also taken by manifoldsR
  uwot_v2 = {
    uwot::umap2(X = benchmark_data$data)
  },
  manifold = {
    umap(
      data = benchmark_data$data,
      nn_method = "balltree", # small dataset, balltree works well...
      k = 15L,
      .verbose = FALSE
    ) 
  },
  times = 1L # single comparison for speed
)
```

Let's pit the two fast versions from uwot and this one against each other:

```{r}
#| label: speed benchmark - large data set
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE) && requireNamespace("umap", quietly = TRUE) && requireNamespace("uwot", quietly = TRUE)

benchmark_data_2 <- manifold_synthetic_data(
  type = "cluster",
  n_samples = 50000L
)

microbenchmark::microbenchmark(
  # uwot - umap2, for example used in Seurat
  uwot_v2 = {
    uwot::umap2(X = benchmark_data_2$data)
  },
  manifold = {
    umap(
      data = benchmark_data_2$data,
      k = 15L,
      .verbose = FALSE
    ) 
  },
  times = 1L # single comparison for speed
)
```

Due to optimised memory layouts, aggressive in-lining, heavy parallelism,
leveragig look-up tables (LUTs), this version is fast and is likely(?) the
fastest UMAP available in R, while keeping memory usage minimal due to 
leveraging Rust frmo the point that data is moved into the Rust kernel, until
the embedding has been fitted.